**The status of the archaeal and bacterial census: an update**

### 2016-03-14 rebuttal.md

**Reviewer #1 (Comments for the Author):**

**The manuscript by Patrick Schloss et al provides an updated on the status of the microbial census. By microbial census the authors refer to the census of bacteria and archaea assessed through the full length 16S rRNA gene sequences.**

**The manuscript is well written and the authors have done a good job in summarizing the current status of the census and the bottlenecks in scaling up.**

**A few minor issues to consider below**
**Page 1. While the manuscript describes a census of bacteria and archaea only, it is referred to as microbial census which however includes microbial eukaryotes. As such the word microbial in the context of this paper and this census is misleading. I assume that the authors would prefer not to use the word prokaryotic census, but microbial is not accurate either.**

**Lines 22-24: The meaning of the sentence "Surprisingly, we have done a relatively poor job of keeping track of the ongoing effort to characterize the biodiversity as represented in full-length 16S rRNA genes", is a bit confusing. There are several databases that keep good track of all the available 16S sequences so the authors should clarify here what they mean.**

**Lines 25-27: The authors claim that "We found that the ongoing effort has done an excellent job of sampling the most abundant organisms, but struggles to sample the more rare organisms". However, recent literature is not very much supportive of that. In a recent publication (Eloe-Fadrosh et al 2016) it is shown that there is a large number of unaffiliated 16S sequences by looking at the assembled metagenomics sequences. Since typically, most of the 16S sequences that get assembled from metagenomic datasets are not of very low abundance organisms, perhaps the statement above doesn't seem to hold up? Furthermore, the authors agree in the discussion that the primers uses for PCR have biases, which contradicts the statement that ongoing efforts have done an excellent job in sampling the most abundant organisms.**

**Lines 142-143: the sentence "Based on the representation of sequences within the SILVA database, in 2006 there were 61 bacterial and 18 phyla" seems to be incomplete. Do they mean 18 "archaeal" phyla?**

**Lines 253-256: The authors discuss the contribution of assembled metagenomics data in showing the presence of introns in 16S rRNA genes. They don't discuss at all the 16S primer mismatches though as another source of significant diversity missed even from abundant organisms as has been recently shown.**

**The authors are also using throughout the text the term "zoonotic" as one of the broad environmental categories. The use of this term is confusing. Do the authors mean host-associated? If yes, I suggest to adopt this term, since this is the term used in the SILVA database and the term recommended from the Genomics Standards Consortium and the MIxS standards. Zoonotic also refers to pathogenic/disease states and unless the authors refer specifically to this type of environments rather than host-associated environments, I would recommend changing the term here.**

**Also, while Table 1 refers to Zoological samples, one of the curves in Figure 1C refers to Zoonotic. I assume that the zoonotic in the figure corresponds to the zoological in the Table, but keeping two terms is confusing, I suggest the authors will also use zoological in the figure to match the Table.**

**It would be also very interesting if the authors would discuss how many of the totally number of OTUs, they generate from the SILVA data remain unaffiliated with any of the known phyla, if any.**

**Reviewer #2 (Comments for the Author):**

**Overall I believe this manuscript is an important and useful extension of the previous microbial census paper. I think the results will be important for the entire community of microbial ecologists and also many others.**

**I have some minor concerns with various statements in the manuscript. They are below**

**Line 2: A census is typically carried out for people at a national level.**
**This is simply not true - censuses are carried out for states, counties, cities, towns, and more**

**Line 6: it would be good in the abstract to mention why only full length sequences are being considered for this census**

**Line 21: I think it would be worth addressing the recent issues in regard to whether the archaea are in fact a separate domain from eukaryotes**

**Line 24: again I think it would be important to explain here why the focus on full length sequences?**

**Line 33: I would argue this is fundamental to more than just ecology**

**Line 42. Regarding the deposition of this sequence in 1983 in Genbank at the NCBI.. I don't think this is possible. I think NCBI did not take over Genbank until later. Plus the paper was in 1978 and the precursor toe Genbank existed before 1983 so it would be good to expain the discrepancies.**

**Line 45: "making it the best-represented gene." details and or/ reference needed for this sentence**

**Line 68-69. "The number of OTUs that are sampled when using different regions". Not completely clear what is meant here - do you mean the # that are inferred from the data or the # that are actually sampled by the PCR for the different regions?**

**Line 74: it would be helpful to clarify what is meant here by "lack the references necessary"**

**Line 92. I think the only thing that can be said from this is saturation for the types of sites being sampled**

**Line 104-105. "it is likely that an even larger number of OTUs have yet to be sampled for both domains." Maybe this is a wording thing but it is unclear to me how we can have "started to saturate" yet have an even larger number of OTUs to be sampled**

**Line 143-145. Since then there have been 4 new bacterial (CKC4, OC31, S2R-29, and SBYG-2791) and 2 new archaeal candidate phyla (Ancient Archaeal Group and TVG8AR30). Citations for this? and is this classification simply pulled fro Silva or based upon other analysis?**

**Line 177. "Great concern" seems excessive. I think "concern expressed" would be more appropriate**

**Line 208. "There is concern". Can you clarify - or, perhaps it would be better to say "we are concerned"**

**Line 233. Problems with EMIRGE. I think this is possibly true. But not enough evidence is presented in this paper to make this type of statement. What evidence do you have that any errors are coming from EMIRGE?**

**Line 292 - 322. Much of the text here discusses data from Silva which appears to have mostly come from EMBL. And other text discusses sequences from other sources. This is all fine. But the text as it is written in the main parts of the paper does not provide enough of this for context. I think it would be very beneficial to add a bit more to the main text on the sources of the sequences.**

**Line 337-338. Figshare. Really good to see what is being released in Figshare and Github for this paper. I would further recommend that sequeneces used in this paper also be included in Figshare so that people do not have to rerun the code in order to get access to the sequences (this is what seems to be required)**

**Line 363. " grow at a rate too slow to ever reach estimates of 10^6 to 10^11 bacterial species." This needs further details in my opinion. What would it take, for example to get to 10^6? How many years at current rates? What if the rate went up a bit? Then how many years?**

**Line 388 "that were also detected by other." I assume some word is missing here.**

### 2016-03-14 submission/Schloss_Census2_mBio_2016.Rmd
Pat's first whack at revisions

---
title: "**The status of the microbial census: an update**"
bibliography: references.bib
output:
  pdf_document:
    includes:
      in_header: header.tex
csl: mbio.csl
fontsize: 11pt
geometry: margin=1.0in
---

@@

Running title: The microbial census

@@

A census is typically carried out for people at a national level; however, microbial ecologists have implemented a molecular census of bacteria and archaea by sequencing their 16S rRNA genes. We assessed how well the microbial census of full-length 16S rRNA gene sequences is proceeding in the context of recent advances in high throughput sequencing technologies. Among the `r coverage_by_category_and_time["total","bact_count"]` and `r coverage_by_category_and_time["total","arch_count"]` full-length bacterial and archaeal sequences sequences, `r coverage_by_category_and_time["total","bact_good_coverage"]`% and `r coverage_by_category_and_time["total","arch_good_coverage"]`% of the bacterial and archeaeal sequences, respectively, belonged to operational taxonomic units (OTUs) that have been observed more than once. Although these metrics suggest that the census is approaching completion, `r coverage_by_category_and_time["total","bact_schloss_coverage"]`% of the bacterial and `r coverage_by_category_and_time["total","arch_schloss_coverage"]`% of the archaeal OTUs have been observed more than once. Thus, there is still considerable microbial diversity to be explored. Unfortunately, the rate of new full-length sequences has been declining and new sequences are primarily being deposited by a small number of studies. Furthermore, sequences from soil and aquatic environments, which are known to be rich in bacterial diversity, only represent `r coarse_percentages["S"]` and `r coarse_percentages["A"]`% of the census while sequences associated with zoonotic environments represent `r coarse_percentages["Z"]`% of the census. Continued use of traditional approaches and new technologies such as single cell genomics and short read assembly are likely to improve our ability to sample rare OTUs if it is possible to overcome this sampling bias. The success of ongoing efforts to use short read sequencing to characterize microbial communities requires that researchers strive to expand the depth and breadth of the microbial census.

@@

The biodiversity contained within the bacterial and archaeal domains dwarfs that of the eukaryotes and the services these organisms provide to the biosphere are critical. Surprisingly, we have done a relatively poor job of keeping track of the ongoing effort to characterize the biodiversity as represented in full-length 16S rRNA genes. By understanding how this census is proceeding, it is possible to suggest the best allocation of resources for advancing the census. We found that the ongoing effort has done an excellent job of sampling the most abundant organisms, but struggles to sample the more rare organisms. Through the use of new sequencing technologies we should be able to obtain full-length sequences from these rare organisms. Furthermore, we suggest that by allocating more resources to sampling environments known to have the greatest biodiversity we will be able to make significant advances in our characterization of microbial diversity.

@@

Generating a comprehensive understanding of any system with a single gene may seem a fool’s errand, yet we have learned a considerable amount regarding the diversity, dynamics, and natural history of microorganisms using the venerable 16S rRNA gene. In 1983, the full-length 16S rRNA gene sequence of *Escherichia coli* (accession J01695) was deposited into NCBI's GenBank making it the first of what is now more than 10 million 16S rRNA gene sequences to be deposited into the database [@Brosius1978]. 16S rRNA gene accessions represent nearly one-third of all sequences deposited in GenBank, making it the best-represented gene. As Sanger sequencing has given way to so-called "next generation sequencing" technologies, hundreds of millions of 16S rRNA gene sequences have been deposited into the NCBI's Sequence Read Archive. The expansion in sequencing throughput and increased access to sequencing technology has allowed for more environments to be sequenced at a deeper coverage, resulting in the identification of novel taxa. The ability to obtain sequence data from microorganisms without cultivation has radically altered our perspective of their role in nearly every environment from deep ocean sediment cores (e.g. accession AY436526) to the International Space Station (e.g. accession DQ497748).

@@

It is disconcerting that the increase in sequencing volume has come at the cost of sequence length. The commonly used MiSeq-based sequencing platform from Illumina is extensively used to sequence the approximately 250 bp V4 hypervariable region of the 16S rRNA gene; other schemes have used different parts of the gene that are generally shorter than 500 bp. The number of OTUs that are sampled when using different regions within the 16S rRNA gene can vary considerably and the genetic diversity within these regions typically has only a modest correlation the genetic diversity of the full-length sequence [@Youssef2009; @Schloss2010]. Thus, it remains unclear to what degree richness estimates from short read technologies over or underestimate the numbers from full-length sequences. Furthermore, we likely lack the references necessary to adequately classify the novel biodiversity we are sampling when we generate 100-times the sequence data from a community than we did using full-length sequencing.

Here we update the status of the microbial census with full-length 16S rRNA gene sequences. In the 13 years since the collection of data for Schloss and Handelsman’s initial census, the number of full-length sequences has grown exponentially, despite the overwhelming contemporary focus by most researchers on short-read technologies. This update to the census allows us to evaluate the relative sampling thoroughness for different environments and clades and make an argument for the continued need to collection full-length sequence data from many systems that have a long history of study. As researchers consider coalescing into a Unified Microbiome Initiative [@Alivisatos2015], it will be important to balance the need for mechanism-based studies with the need to generate full-length reference sequences from a diversity of environments.

@@

***Sequencing efforts are a source of bias in the census.***
One explanation for the large number of OTUs that have only been observed once is that with the the broad adoption of sequencing platforms that generate short sequence reads, the rate of full-length sequence generation has declined. In fact, since 2009 the number of new bacterial sequences generated has slowed to an average of `r floor(mean(time_data[as.character(2009:2014),'b_nseqs']))` sequences per year (Figure 2A). Although this is still an impressive number of sequences, since `r start<-2007;format(start, big.mark='')` the number of new bacterial OTUs has plateaued at an average of `r round(mean(time_data[as.character(start:2014),'b_sobs']),0)` new OTUs per year (Figure 2B). Given the expense of generating full-length sequences using the Sanger sequencing technology and the transition to other platforms at that time, we expected that the large number of sequences were being deposited by a handful of large projects. Indeed, when we counted the number of submissions responsible for depositing 50% of the sequences, we found that with the exception of 2006 and 2013, eight or fewer studies were responsible for depositing the majority of the full-length sequences each year since 2005 (Figure 2C). Between 2009 and 2012, `r all_nseqs_2009_2012` total sequences were submitted and `r sum(time_data[as.character(2009:2012), 'b_n50'])` submissions from `r length(dominant_pmids_2009_2012)` studies were responsible for depositing `r dominant_nseqs_2009_2012` (`r 100*dominant_nseqs_2009_2012/all_nseqs_2009_2012`% of all sequences). These studies generated sequences from the human gastrointestinal tract [@Li_2012], human skin [@Kong_2012; @Grice_2009], murine skin [@Grice_2010], and hypersaline microbial mats [@Kirk_Harris_2012]. The heavy zoonotic focus is reflected in the rarefaction curve for this category (Figure 1C). In contrast to recent years, between 1995 and 2006, an average of `r mean(time_data[as.character(1995:2006), 'b_n50'])` studies were responsible for submitting more than half of the sequences each year. Although the recent deep surveys represent significant contributions to our knowledge of bacterial biogeography, their small number and lack of environmental diversity is indicative of the broader problems in advancing the bacterial census.

@@

***The ability to sample microbial life is taxonomically skewed.***
The Firmicutes, Proteobacteria, Actinobacteria, and Bacteroidetes represent `r 100*sum(b_phylum[c("Firmicutes", "Proteobacteria", "Actinobacteria", "Bacteroidetes"), "all"])/sum(b_phylum$all)`% of the bacterial sequences and the Euryarchaeota and Thaumarchaeota `r 100*sum(a_phylum[c("Euryarchaeota",  "Thaumarchaeota"), "all"])/sum(a_phylum$all)`% of the archaeal sequences. We sought to understand how the representation of individual phyla has changed relative to the state of the census in 2006. We used 2006 as a reference point for calibrating the dynamics of the bacterial and archaeal censuses since that was the year that the first highly parallelized 16S rRNA gene sequence dataset was published [@Sogin2006]. Based on the representation of sequences within the SILVA database, in 2006 there were `r sum(b_phylum$pre!=0)` bacterial and `r sum(a_phylum$pre!=0)` phyla. Since then there have been `r length(b_new_phyla)` new  bacterial (`r b_new_phyla_names`) and `r length(a_new_phyla)` new archaeal candidate phyla (`r a_new_phyla_names`). Relative to the overall sequencing trends before and after 2006, several phyla stand out for being over and underrepresented in sequence submissions (Figure 3). Among the bacterial phyla with at least 1,000 sequences, `r oxford_comma(sort(rownames(b_phylum[b_phylum$ratio_of_ratios >= 4.0 & b_phylum$all >=1000,])))` were sequenced 4-fold more often while `r oxford_comma(sort(rownames(b_phylum[b_phylum$ratio_of_ratios < 0.5 & b_phylum$all >=1000,])))` were sequenced 2-fold less often than would have been expected since 2006. Among the archaeal phyla with at least 1,000 sequences, the Thaumarchaeota were sequenced `r a_phylum["Thaumarchaeota", "ratio_of_ratios"]`-fold more often and the Crenarchaeota were sequenced `r 1/a_phylum["Crenarchaeota", "ratio_of_ratios"]`-fold less often than expected. Together, these results demonstrate a change in the phylum-level lineages represented in the census from before and after 2006 and encouragingly, show that some underrepresented phyla are becoming better sampled.

@@

***The cultured census.***
In the 2004 bacterial census, there was great concern that although culture-independent methods were significantly enhancing our knowledge of microbial life, there were numerous bacterial phyla with no or only a few cultured representatives. To update this assessment, we identified those sequences that came from cultured and uncultured organisms. Overall, `r b_percent_cult_nseqs_all`% of bacterial sequences and `r a_percent_cult_nseqs_all`% of archaeal sequences have come from isolated organisms. Comparing the fraction of sequences deposited during and before 2006 from isolates to those collected after 2006, we found that culturing rates lag by `r  b_shift_cult_nseqs_all` and `r a_shift_cult_nseqs_all`-fold for bacteria and archaea, respectively. Among the `r nrow(b_cult_time)` bacterial phyla, `r sum(b_cult_time$all_cult_nseqs == 0)` have no cultured representatives and `r sum(a_cult_time$all_cult_nseqs == 0)` of the `r nrow(a_cult_time)`  archaeal phyla have no cultured representatives. This lag is likely due to the differences in throughput of culture-dependent and -independent approaches. Of the phyla with at least one cultured representative, the median percentage of sequences coming from a culture was only `r median(b_rate_cult_nseqs_phyulm[b_rate_cult_nseqs_phyulm!=0])*100`% for the bacterial phyla and `r median(a_rate_cult_nseqs_phyulm[a_rate_cult_nseqs_phyulm!=0])*100`% for the archaeal phyla (Figure 5). Even though many phyla have cultured representatives, there is still a skew in the representation of most phyla found in cultivation efforts.

@@

Given the shift from Sanger sequencing to platforms that offer higher throughput but shorter reads, there is concern that our ability to harvest full-length sequences from communities will remain stalled. Several culture-independent methods have been developed that offer the ability to obtain full-length sequences of the 16S rRNA gene and even the complete genome. These have included single cell genomics [@Rinke2013] and assembly of short 16S rRNA gene fragments using data generated from PCR amplicons or metagenomic shotgun sequence data with the Expectation-Maximization Iterative Reconstruction of Genes from the Environment (EMIRGE) algorithm [@Miller2011; @Miller2013]. To test the ability of these technologies to expand our knowledge of microbial diversity beyond that of traditional approaches, we compared the overlap of OTUs found using each of the new methods with the traditional approaches (Figure 6). Utilizing the 16S rRNA gene sequences extracted from the single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@Markowitz2013], we identified `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'n_seqs']` bacterial and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'n_seqs']` archaeal sequences, which were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'n_otus']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'n_otus']` bacterial and archaeal OTUs, respectively. Interestingly, only `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'sc']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'sc']`% of the bacterial and archaeal single celled OTUs, respectively, had not been observed by previous efforts. Next, we identified six studies that utilized EMIRGE to assemble 16S rRNA gene sequences from metagenomic sequences [@Wrighton2012; @Miller2011; @NunesdaRocha2015; @Hamilton2015; @Handley2012; @Gladden2011]. Together these studies assembled `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'n_seqs']` bacterial and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'n_seqs']` archaeal full-length sequences, which were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'n_otus']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'n_otus']` bacterial and archaeal OTUs, respectively. Only

@@

There are several additional reasons to suspect that the current census should be considered conservative. First, we found that most sequences recently deposited into public databases are being made by a small number of projects that have deeply sampled similar environments, and the number of full-length reads deposited into the databases has stalled. Second, it is widely acknowledged that 16S rRNA gene primers are biased; these biases are amplified when designing primers to amplify subregions used in sequencing short reads [@Parada2015]. Assembly of metagenomic data has shown the presence of introns in the 16S rRNA genes of organisms within the so-called "Candidate Phyla Radiation” (e.g. Saccharibacteria (TM7), Peregrinibacteria, Berkelbacteria (ACD58), WWE3 Microgenomates (OP11), Parcubacteria (OD1), et al.) that would preclude detection with standard PCR-based approaches [@Brown2015; @EloeFadrosh2016]. Third, the willingness of researchers to contribute their sequences and the metadata describing the environment that the sequences were sampled from is critical for assessing the progress of the census and to accrue the benefits from having full-length sequences in the databases. Interestingly, the first 16S sequence was published in 1978, but was not available in a database until 1983. Similarly, only 5 of the 11 studies that used the EMIRGE algorithm deposited their sequences in GenBank. This makes the sequences from the other studies effectively invisible to the search algorithms used by 16S rRNA gene-specific databases to harvest sequences. As assembly and long read technologies advance, a mechanism is needed to assess the quality of the consensus sequences and to make them easily accessible to the 16S rRNA gene-specific databases.

>>>>

---
title: "**The status of the archaeal and bacterial census: an update**"
bibliography: references.bib
output:
  pdf_document:
    keep_tex: true
    includes:
      in_header: header.tex
csl: mbio.csl
fontsize: 11pt
geometry: margin=1.0in
---

@@

Running title: The archaeal and bacterial census

@@

A census is typically carried out for people across a range of geographical levels; however, microbial ecologists have implemented a molecular census of bacteria and archaea by sequencing their 16S rRNA genes. We assessed how well the census of full-length 16S rRNA gene sequences is proceeding in the context of recent advances in high throughput sequencing technologies because full-length sequences are typically used as references for classification of the short sequences generated by newer technologies. Among the `r coverage_by_category_and_time["total","bact_count"]` and `r coverage_by_category_and_time["total","arch_count"]` full-length bacterial and archaeal sequences sequences, `r coverage_by_category_and_time["total","bact_good_coverage"]`% and `r coverage_by_category_and_time["total","arch_good_coverage"]`% of the bacterial and archeaeal sequences, respectively, belonged to operational taxonomic units (OTUs) that have been observed more than once. Although these metrics suggest that the census is approaching completion, `r coverage_by_category_and_time["total","bact_schloss_coverage"]`% of the bacterial and `r coverage_by_category_and_time["total","arch_schloss_coverage"]`% of the archaeal OTUs have been observed more than once. Thus, there is still considerable diversity to be explored. Unfortunately, the rate of new full-length sequences has been declining and new sequences are primarily being deposited by a small number of studies. Furthermore, sequences from soil and aquatic environments, which are known to be rich in bacterial diversity, only represent `r coarse_percentages["S"]` and `r coarse_percentages["A"]`% of the census while sequences associated with host-associated environments represent `r coarse_percentages["Z"]`% of the census. Continued use of traditional approaches and new technologies such as single cell genomics and short read assembly are likely to improve our ability to sample rare OTUs if it is possible to overcome this sampling bias. The success of ongoing efforts to use short read sequencing to characterize archaeal and bacterial communities requires that researchers strive to expand the depth and breadth of this census.

@@

The biodiversity contained within the bacterial and archaeal domains dwarfs that of the eukaryotes and the services these organisms provide to the biosphere are critical. Surprisingly, we have done a relatively poor job of formally tracking  the quality of the biodiversity as represented in full-length 16S rRNA genes. By understanding how this census is proceeding, it is possible to suggest the best allocation of resources for advancing the census. We found that the ongoing effort has done an excellent job of sampling the most abundant organisms, but struggles to sample the more rare organisms. Through the use of new sequencing technologies we should be able to obtain full-length sequences from these rare organisms. Furthermore, we suggest that by allocating more resources to sampling environments known to have the greatest biodiversity we will be able to make significant advances in our characterization of archaeal and bacterial diversity.

@@

Generating a comprehensive understanding of any system with a single gene may seem a fool’s errand, yet we have learned a considerable amount regarding the diversity, dynamics, and natural history of microorganisms using the venerable 16S rRNA gene. In 1983, the full-length 16S rRNA gene sequence of *Escherichia coli* (accession J01695) was released as part of NCBI's GenBank making it the first of what is now more than 10 million 16S rRNA gene sequences to be deposited into the database [@Brosius1978]. 16S rRNA gene accessions represent nearly one-third of all sequences deposited in GenBank, making it the best-represented gene. As Sanger sequencing has given way to so-called "next generation sequencing" technologies, hundreds of millions of 16S rRNA gene sequences have been deposited into the NCBI's Sequence Read Archive. The expansion in sequencing throughput and increased access to sequencing technology has allowed for more environments to be sequenced at a deeper coverage, resulting in the identification of novel taxa. The ability to obtain sequence data from microorganisms without cultivation has radically altered our perspective of their role in nearly every environment from deep ocean sediment cores (e.g. accession AY436526) to the International Space Station (e.g. accession DQ497748).

@@

It is disconcerting that the increase in sequencing volume has come at the cost of sequence length. The commonly used MiSeq-based sequencing platform from Illumina is extensively used to sequence the approximately 250 bp V4 hypervariable region of the 16S rRNA gene; other schemes have used different parts of the gene that are generally shorter than 500 bp. The number of OTUs that are inferred from the sequencing data when using different regions within the 16S rRNA gene can vary considerably and the genetic diversity within these regions typically has only a modest correlation with the genetic diversity of the full-length sequence [@Youssef2009; @Schloss2010]. Thus, it remains unclear to what degree richness estimates from short read technologies over or underestimate the numbers from full-length sequences. Furthermore, we likely lack the full-length reference sequences necessary to adequately classify the novel biodiversity we are sampling when we generate 100-times the sequence data from a community than we did using full-length sequencing.

Here we update the status of the archaeal and bacterial census with full-length 16S rRNA gene sequences. In the 13 years since the collection of data for Schloss and Handelsman’s initial census, the number of full-length sequences has grown exponentially, despite the overwhelming contemporary focus by most researchers on short-read technologies. This update to the census allows us to evaluate the relative sampling thoroughness for different environments and clades and make an argument for the continued need to collection full-length sequence data from many systems that have a long history of study. As researchers consider coalescing into a Unified Microbiome Initiative [@Alivisatos2015], it will be important to balance the need for mechanism-based studies with the need to generate full-length reference sequences from a diversity of environments.

@@

***Sequencing efforts are a source of bias in the census.***
One explanation for the large number of OTUs that have only been observed once is that with the the broad adoption of sequencing platforms that generate short sequence reads, the rate of full-length sequence generation has declined. In fact, since 2009 the number of new bacterial sequences generated has slowed to an average of `r floor(mean(time_data[as.character(2009:2014),'b_nseqs']))` sequences per year (Figure 2A). Although this is still an impressive number of sequences, since `r start<-2007;format(start, big.mark='')` the number of new bacterial OTUs has plateaued at an average of `r round(mean(time_data[as.character(start:2014),'b_sobs']),0)` new OTUs per year (Figure 2B). Given the expense of generating full-length sequences using the Sanger sequencing technology and the transition to other platforms at that time, we expected that the large number of sequences were being deposited by a handful of large projects. Indeed, when we counted the number of submissions responsible for depositing 50% of the sequences, we found that with the exception of 2006 and 2013, eight or fewer studies were responsible for depositing the majority of the full-length sequences each year since 2005 (Figure 2C). Between 2009 and 2012, `r all_nseqs_2009_2012` total sequences were submitted and `r sum(time_data[as.character(2009:2012), 'b_n50'])` submissions from `r length(dominant_pmids_2009_2012)` studies were responsible for depositing `r dominant_nseqs_2009_2012` (`r 100*dominant_nseqs_2009_2012/all_nseqs_2009_2012`% of all sequences). These studies generated sequences from the human gastrointestinal tract [@Li_2012], human skin [@Kong_2012; @Grice_2009], murine skin [@Grice_2010], and hypersaline microbial mats [@Kirk_Harris_2012]. The heavy focus on host-associated communities is reflected in the rarefaction curve for this category (Figure 1C). In contrast to recent years, between 1995 and 2006, an average of `r mean(time_data[as.character(1995:2006), 'b_n50'])` studies were responsible for submitting more than half of the sequences each year. Although the recent deep surveys represent significant contributions to our knowledge of bacterial biogeography, their small number and lack of environmental diversity is indicative of the broader problems in advancing the bacterial census.

@@

***The ability to sample archaea and bacteria is taxonomically skewed.***
The Firmicutes, Proteobacteria, Actinobacteria, and Bacteroidetes represent `r 100*sum(b_phylum[c("Firmicutes", "Proteobacteria", "Actinobacteria", "Bacteroidetes"), "all"])/sum(b_phylum$all)`% of the bacterial sequences and the Euryarchaeota and Thaumarchaeota `r 100*sum(a_phylum[c("Euryarchaeota",  "Thaumarchaeota"), "all"])/sum(a_phylum$all)`% of the archaeal sequences. We sought to understand how the representation of individual phyla has changed relative to the state of the census in 2006. We used 2006 as a reference point for calibrating the dynamics of the bacterial and archaeal censuses since that was the year that the first highly parallelized 16S rRNA gene sequence dataset was published [@Sogin2006]. Based on the representation of sequences within the SILVA database, in 2006 there were `r sum(b_phylum$pre!=0)` bacterial and `r sum(a_phylum$pre!=0)` archaeal phyla. Since then there have been `r length(b_new_phyla)` new  bacterial (`r b_new_phyla_names`) and `r length(a_new_phyla)` new archaeal candidate phyla (`r a_new_phyla_names`). Relative to the overall sequencing trends before and after 2006, several phyla stand out for being over and underrepresented in sequence submissions (Figure 3). Among the bacterial phyla with at least 1,000 sequences, `r oxford_comma(sort(rownames(b_phylum[b_phylum$ratio_of_ratios >= 4.0 & b_phylum$all >=1000,])))` were sequenced 4-fold more often while `r oxford_comma(sort(rownames(b_phylum[b_phylum$ratio_of_ratios < 0.5 & b_phylum$all >=1000,])))` were sequenced 2-fold less often than would have been expected since 2006. Among the archaeal phyla with at least 1,000 sequences, the Thaumarchaeota were sequenced `r a_phylum["Thaumarchaeota", "ratio_of_ratios"]`-fold more often and the Crenarchaeota were sequenced `r 1/a_phylum["Crenarchaeota", "ratio_of_ratios"]`-fold less often than expected. Together, these results demonstrate a change in the phylum-level lineages represented in the census from before and after 2006 and encouragingly, show that some underrepresented phyla are becoming better sampled.

@@

***The cultured census.***
In the 2004 bacterial census, there was concern expressed that although culture-independent methods were significantly enhancing our knowledge of microbial life, there were numerous bacterial phyla with no or only a few cultured representatives. To update this assessment, we identified those sequences that came from cultured and uncultured organisms. Overall, `r b_percent_cult_nseqs_all`% of bacterial sequences and `r a_percent_cult_nseqs_all`% of archaeal sequences have come from isolated organisms. Comparing the fraction of sequences deposited during and before 2006 from isolates to those collected after 2006, we found that culturing rates lag by `r  b_shift_cult_nseqs_all` and `r a_shift_cult_nseqs_all`-fold for bacteria and archaea, respectively. Among the `r nrow(b_cult_time)` bacterial phyla, `r sum(b_cult_time$all_cult_nseqs == 0)` have no cultured representatives and `r sum(a_cult_time$all_cult_nseqs == 0)` of the `r nrow(a_cult_time)`  archaeal phyla have no cultured representatives. This lag is likely due to the differences in throughput of culture-dependent and -independent approaches. Of the phyla with at least one cultured representative, the median percentage of sequences coming from a culture was only `r median(b_rate_cult_nseqs_phyulm[b_rate_cult_nseqs_phyulm!=0])*100`% for the bacterial phyla and `r median(a_rate_cult_nseqs_phyulm[a_rate_cult_nseqs_phyulm!=0])*100`% for the archaeal phyla (Figure 5). Even though many phyla have cultured representatives, there is still a skew in the representation of most phyla found in cultivation efforts.

@@

Given the shift from Sanger sequencing to platforms that offer higher throughput but shorter reads, we are concerned that our ability to harvest full-length sequences from communities will remain stalled. Several culture-independent methods have been developed that offer the ability to obtain full-length sequences of the 16S rRNA gene and even the complete genome. These have included single cell genomics [@Rinke2013] and assembly of short 16S rRNA gene fragments using data generated from PCR amplicons or metagenomic shotgun sequence data with the Expectation-Maximization Iterative Reconstruction of Genes from the Environment (EMIRGE) algorithm [@Miller2011; @Miller2013]. To test the ability of these technologies to expand our knowledge of microbial diversity beyond that of traditional approaches, we compared the overlap of OTUs found using each of the new methods with the traditional approaches (Figure 6). Utilizing the 16S rRNA gene sequences extracted from the single-cell genomes available on the Integrated Microbial Genomes (IMG) system [@Markowitz2013], we identified `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'n_seqs']` bacterial and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'n_seqs']` archaeal sequences, which were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'n_otus']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'n_otus']` bacterial and archaeal OTUs, respectively. Interestingly, only `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="sc", 'sc']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="sc", 'sc']`% of the bacterial and archaeal single celled OTUs, respectively, had not been observed by previous efforts. Next, we identified six studies that utilized EMIRGE to assemble 16S rRNA gene sequences from metagenomic sequences [@Wrighton2012; @Miller2011; @NunesdaRocha2015; @Hamilton2015; @Handley2012; @Gladden2011]. Together these studies assembled `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'n_seqs']` bacterial and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'n_seqs']` archaeal full-length sequences, which were assigned to `r overlap_counts[overlap_counts$domain=="bacteria" & overlap_counts$method=="em_meta", 'n_otus']` and `r overlap_counts[overlap_counts$domain=="archaea" & overlap_counts$method=="em_meta", 'n_otus']` bacterial and archaeal OTUs, respectively. Only

@@

There are several additional reasons to suspect that the current census should be considered conservative. First, we found that most sequences recently deposited into public databases are being made by a small number of projects that have deeply sampled similar environments, and the number of full-length reads deposited into the databases has stalled. Second, it is widely acknowledged that 16S rRNA gene primers are biased; these biases are amplified when designing primers to amplify subregions used in sequencing short reads [@Parada2015]. Assembly of metagenomic data has shown the presence of introns in the 16S rRNA genes of organisms within the so-called "Candidate Phyla Radiation” (e.g. Saccharibacteria (TM7), Peregrinibacteria, Berkelbacteria (ACD58), WWE3 Microgenomates (OP11), Parcubacteria (OD1), et al.) that would preclude detection with standard PCR-based approaches [@Brown2015; @EloeFadrosh2016]. Third, the willingness of researchers to contribute their sequences and the metadata describing the environment that the sequences were sampled from is critical for assessing the progress of the census and to accrue the benefits from having full-length sequences in the databases. As an illustration of this problem, only 5 of the 11 studies that used the EMIRGE algorithm deposited their sequences in GenBank. This makes the sequences from the other studies effectively invisible to the search algorithms used by 16S rRNA gene-specific databases to harvest sequences. As assembly and long read technologies advance, a mechanism is needed to assess the quality of the consensus sequences and to make them easily accessible to the 16S rRNA gene-specific databases.

Efforts to census archaea and bacteria using short read technology such as the International Census of Marine Microbes, the Earth Microbiome Project, and the Human Microbiome Project have significantly advanced our knowledge of archaeal and bacterial biogeography; however, these analyses have demonstrated the limitations of databases and taxonomies that are based on sequences from common and abundant organisms. During the period prior to the introduction of massively parallelized high throughput sequencing, it was common for a study to generate dozens or hundreds of sequences per sample. The existing databases that are used for classifying sequences are based on those sequences, which represent organisms that are generally abundant. We hypothesize that recent difficulties obtaining adequate classification for short sequences captured from more rare organisms are because our databases do not contain full-length references for those sequences. We fear that these trends will worsen unless researchers can leverage new sequencing and cultivation technologies to generate large numbers of full-length sequences from a large number of diverse samples.
